A molecular dynamics study of the solvation of carbon dioxide in the ionic liquids \ce{[emim][B(CN)_4]} and \ce{[emim][NTf_2]}

### 2018-09-04 ionic_liquids.tex
Reformulation of the Solvation Free Energy subsection

The solvation free energy, $\Delta G_{\text{solv}}$, is defined as the reversible work needed to transfer a solute molecule from a gas phase to the solution.
For dense systems, an efficient computation of solvation properties via molecular simulation is possible by means of the so-called alchemical methods, which employ non-physical pathways to reversibly connect the endpoint states whose difference in free energy is to be estimated.
In the case of $\Delta G_{\text{solv}}$, the alchemical route involves a gradual insertion/deletion of the solute through the use of a coupling parameter $\lambda$, which scales the solute-solvent interactions by taking values in the interval [0,1]. By convention, $\lambda = 1$ corresponds to a fully interacting system, while for $\lambda$ = 0 the solute-solvent interactions are turned off completely .

In what follows, the subscripts ``s'' and ``IL'' refer to the solute and the solvent (ionic liquid), respectively.
For a system containing $N_\text{\tiny IL}$ solvent molecules and $N_{\text{s}}$ solute molecules at fixed pressure (P) and temperature (T), the difference in free energy between the states associated to $\lambda$ = 1 and $\lambda$ = 0 is formally defined as
\begin{equation}
\Delta G_\text{sim} \equiv k_b T \ln \frac{\bigtriangleup (N_s^l,N_\text{\tiny IL},P,T,\lambda = 1)}{\bigtriangleup (N_s^l,N_\text{\tiny IL},P,T,\lambda = 0)}, 
\end{equation}
where superscript ``$l$'' refers to the liquid solution, $\bigtriangleup$ denotes the isobaric-isothermal partition function, and $k_b$ is the Boltzmann's constant.
Shirts \textit{et al.} \cite{Shirts_2003} rigorously derived the working expression that relates $\Delta G_\text{sim}$ to what is experimentally known as the solvation free energy.
They found out that
\begin{equation}
\begin{split}
\label{eq:free_solv}
 \Delta G_{\text{solv}}& =  k_b T \ln \left( \frac{P_s}{c_s^{\ast} k_b T} \right)\\ &= \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1} \right),
\end{split}
\end{equation}
where $V^{\ast}$ and $V_1$ refer to the mean volumes of the systems specified by ($N_s^l-1$,$N_\text{\tiny IL}$,P,T) and ($N_s^l$,$N_\text{\tiny IL}$,P,T), respectively; $c_s^{\ast}$ = $N_s^l/V_1$ is the numerical concentration of the solute in the liquid-phase and $P_s$ is the partial pressure of the solute in the gas phase.
It should be noted that in the derivation of Eq.~\eqref{eq:free_solv}, the gas phase is treated as an ideal gas.

For small solutes, such as those considered in this work, the magnitude of the term $V^{\ast}/V_1$ is typically on the order of the free energy uncertainties \cite{Shirts_2003}.
Hence, that term can be neglected and will be omitted hereafter.

\subsection*{Henry constant}
If both $P_s$ and $c_s^{\ast}$ were not large, we can resort to Eq.~\eqref{eq:free_solv} to predict Henry's law constant $\text{K}_s$, which is defined as \cite{Prausnitz}
\begin{equation}
\text{K}_s =\frac{P_s}{x_s}.
\end{equation}
Therefore, by simply expressing Eq.~\eqref{eq:free_solv} in terms of the molar fraction $x_s$ and assuming that MW (molecular weight of the solution) = $\text{MW}_\text{\tiny IL}$, as well as $\rho$ (mass density of the solution) = $\rho_\text{\tiny IL}$, we obtain that
\begin{equation}
\label{eq:henry_eq}
K_s = \tilde{\rho}_\text{\tiny IL} R \, T \exp(\beta \, \Delta G_\text{sim}),
\end{equation}
where $\tilde{\rho}_\text{\tiny IL}$ is the molar density of the solvent and $R$ is the gas constant.

@@

For the ILs of interest to the present investigation, high concentrations of \ce{CO_2} are found at pressure values for which, in principle, the ideal gas assumption may be not valid.
In this case, we shall correct the ideal gas contribution to the chemical potential, as follows
\begin{equation}
\label{eq:mu_gas_real}
\mu^{g}_s = \mu^{ig}_s + RT \ln \phi_s,
\end{equation}
where $\phi_s$ is the fugacity coefficient of the solute.
This equation, together with the fact that $f^{g}_s$ = $P_s \phi_s$, are used to evaluate $\mu^{l}_s$ = $\mu^{g}_s$, which yields
\begin{equation}
\label{eq:fgas_d}
f^{g}_s = \exp  \left( \beta \Delta G_\text{sim} \right) +  k_b T c_s^{\ast}, 
\end{equation}
where $f^{g}_s$ is the fugacity of the solute in the gas phase.
Having calculated $f^{g}_s$, $P_s$ can be predicted using the following equation
\begin{equation}
\label{eq:fgas}
f^{g}_s = P_s \exp\left[ \frac{B_s(T) P_s}{R T} \right],
\end{equation}
where $B_s$ corresponds to the second virial coefficient.
Here, we employ the correlation of $B_s$ for \ce{CO_2} introduced by Holste \textit{et al.} \cite{Holste_1987}, which is
\begin{equation}
B_s = B_0 + \frac{B_1}{T} + \frac{B_2}{T^2},
\end{equation}
where $B_0$ = 23.02991 cm$^3$/mol, $B_1$ = -2.455297 $\times$ 10$^3$ (Kcm$^3$/mol) and $B_2$ = -1.22675$\times$10$^7$ (K$^2$cm$^3$/mol).

>>>>

The solvation free energy, $\Delta G_{\text{solv}}$, is defined as the reversible work needed to transfer a solute molecule from a gas phase to the solution.
For dense systems, an efficient computation of solvation properties via molecular simulation is possible by means of the so-called alchemical methods, which employ non-physical pathways to reversibly connect the endpoint states whose difference in free energy is to be estimated.
In the case of $\Delta G_{\text{solv}}$, the alchemical route involves a gradual insertion/deletion of the solute through the use of a coupling parameter $\lambda$, which scales the solute-solvent interactions by taking values in the interval [0,1]. By convention, $\lambda = 1$ corresponds to a fully interacting system, while for $\lambda$ = 0 the solute-solvent interactions are turned off completely.

In what follows, the subscripts ``s'' and ``IL'' refer to the solute and the solvent (ionic liquid), respectively.
For a liquid phase containing $N_\text{\tiny IL}$ solvent molecules and $N_{\text{s}}$ solute molecules at fixed pressure (P) and temperature (T), the difference in free energy between the states associated to $\lambda$ = 1 and $\lambda$ = 0 is formally defined as
\begin{equation}
\Delta G_\text{sim} \equiv k_b T \ln \frac{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 1)}{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 0)}, 
\end{equation}
where $\Xi$ denotes the isobaric-isothermal partition function and $k_b$ is the Boltzmann constant.
Shirts \textit{et al}. \cite{Shirts_2003} rigorously derived the working expression that relates $\Delta G_\text{sim}$ to what is experimentally measured as the solvation free energy, namely
\begin{equation}
\Delta G_{\text{solv}} =  k_b T \ln \left( \frac{P_s}{c^l_s R T} \right),
\end{equation}
where $R=N_Ak_b$ is the gas constant ($N_A$ is the Avogadro number) and, for a system at phase equilibrium, $P_s$ is the partial pressure of the solute in the gas phase and $c^l_s$ is its molar concentration in the liquid phase. For obtaining the mentioned relation, the authors derived an expression for the chemical potential of the solute in the liquid phase ($\mu^l_s$) as a function of $\Delta G_\text{sim}$, which can be written as
\begin{equation}
\mu^l_s = \mu^{ig}_s + \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1}\right) - k_bT \ln \left( \frac{P_s}{c_s^{\ast} k_b T} \right),
\end{equation}
where $\mu^{ig}_s$ is the chemical potential of the solute in an ideal gas at temperature $T$ and solute partial pressure $P_s$, $V^{\ast}$ and $V_1$ are the mean volumes of liquid solutions specified by ($N_s-1$,$N_\text{\tiny IL}$,P,T) and ($N_s$,$N_\text{\tiny IL}$,P,T), respectively, while $c_s^{\ast} = N_s/V_1$ is the number concentration of the solute in the latter.
For small solutes, such as those considered in this work, the term containing the ratio $V^{\ast}/V_1$ is typically in the order of magnitude of the free-energy uncertainties \cite{Shirts_2003}.
Hence, it can be neglected and will be omitted hereafter.

Finally, by treating the gas phase as an ideal gas and considering a system at phase equilibrium ($\mu^l_s = \mu^{ig}_s$), Shirts \textit{et al}. \cite{Shirts_2003} then found out that
\begin{equation}
\begin{split}
\label{eq:free_solv}
 \Delta G_{\text{solv}} \approx \Delta G_\text{sim}.
\end{split}
\end{equation}

\subsection*{Henry constant}

If both $P_s$ and $c_s^{\ast}$ are small, we can resort to Eq.~\eqref{eq:free_solv} to predict Henry's law constant $\text{K}_s$, which is defined as \cite{Prausnitz}
\begin{equation}
\text{K}_s =\frac{P_s}{x_s}.
\end{equation}

Therefore, by simply expressing Eq.~\eqref{eq:free_solv} in terms of the molar fraction $x_s$ and assuming that the molar density of the solution does not perceivably differ from than of the solvent ($\tilde{\rho}_\text{\tiny solution} \approx \tilde{\rho}_\text{\tiny IL}$), we obtain that
\begin{equation}
\label{eq:henry_eq}
K_s = \tilde{\rho}_\text{\tiny IL} R \, T \exp(\beta \, \Delta G_\text{sim}),
\end{equation}
where $R$ is the gas constant and $\beta = (k_b T)^{-1}$.

@@

For the ILs of interest to the present investigation, high concentrations of \ce{CO_2} are found at pressure values for which, in principle, the ideal gas assumption may be not valid.
In this case, we can compute the chemical potential of the solute in the real gas phase as
\begin{equation}
\label{eq:mu_gas_real}
\mu^{g}_s = \mu^{ig}_s + RT \ln \phi_s,
\end{equation}
where $\phi_s$ is the fugacity coefficient of the solute.
This equation, together with the fact that $f^{g}_s$ = $P_s \phi_s$, are used to evaluate the phase equilibrium equation $\mu^{l}_s$ = $\mu^{g}_s$, which yields
\begin{equation}
\label{eq:fgas_d}
f^{g}_s = \exp  \left( \beta \Delta G_\text{sim} \right) +  k_b T c_s^{\ast}, 
\end{equation}
where $f^{g}_s$ is the fugacity of the solute in the gas phase.
Having calculated $f^{g}_s$, $P_s$ can be predicted using the following equation
\begin{equation}
\label{eq:fgas}
f^{g}_s = P_s \exp\left[ \frac{B_s(T) P_s}{R T} \right],
\end{equation}
where $B_s$ corresponds to the second virial coefficient.
Here, we employ the correlation of $B_s$ for \ce{CO_2} introduced by Holste \textit{et al.} \cite{Holste_1987}, which is
\begin{equation}
B_s = B_0 + \frac{B_1}{T} + \frac{B_2}{T^2},
\end{equation}
where $B_0$ = 23.02991 cm$^3$/mol, $B_1$ = -2.455297 $\times$ 10$^3$ (Kcm$^3$/mol) and $B_2$ = -1.22675$\times$10$^7$ (K$^2$cm$^3$/mol).
